---
title: "Case study: differential gene expression analysis"
author: "Nicolas Enjalbert Courrech and Pierre Neuvial"
date: today
format:
  html:
    toc: true
    number-sections: true
  pdf:
    number-sections: true
bibliography: sanssouci.bib
execute:
  cache: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  cache = TRUE
)
```

This vignette illustrates the use of `quarto` to build a reproducible data analysis report. 

```{r}
#| label: install-r-packages
#| results: hide
#| message: false
#| eval: false
require("ggplot2") || install.packages("ggplot2")
require("sanssouci") || remotes::install_github("sanssouci-org/sanssouci@develop")
```

```{r}
#| label: load-r-packages
#| results: hide
#| message: false
library("ggplot2")
library("sanssouci")
library("matrixStats")
```

# Motivation: a differential gene expression study

Differential gene expression studies aim at identifying genes whose mean expression level differs significantly between two (or more) populations, based on a sample of gene expression measurements from individuals from these populations. The data can be loaded as follows:

```{r}
#| label: load-data
#| message: false
data("RNAseq_blca", package = "sanssouci.data")
Y <- RNAseq_blca
groups <- ifelse(colnames(RNAseq_blca) == "III", 1, 0)
rm(RNAseq_blca)
```

This data set consists of gene expression measurements for  $n = `r ncol(Y)`$ patients from the Cancer Genome Atlas Urothelial Bladder Carcinoma (TCGA-BLCA) data collection (@tcga2014blca). Following standard practice in RNAseq data analysis, we convert these measurements to to *count per millions* (CPM) in order to make them comparable across samples.

```{r}
#| label: fig-CPM
#| fig-cap: "Empirical distribution of gene expression across all genes and samples"
CPM <- Y/colSums(Y)*1e6
plot(density(log(1 + CPM)))
```

We now filter out genes with very low counts in all samples and plot the corresponding distribution:

```{r}
#| label: fig-CPM-filter
#| fig-cap: "Empirical distribution of gene expression across all samples after gene filtering"
ww <- which(rowQuantiles(log(1 + CPM), prob = 0.75) < log(1 + 5))
Y <- CPM[-ww, ]
plot(density(log(1 + Y)))
m <- nrow(Y)
```

The first mode in @fig-CPM has been filtered out in @fig-CPM-filter.
The patients are classified into two subgoups, according to their clinical stage (II or III):

```{r}
#| label: colnames
table(groups)
```

The goal of this study is to understand the molecular differences at the gene expression level between the stage II and stage III populations. Each patient is characterized by a vector of $m = `r m`$ gene expression values.  We aim at addressing the following question:

> For which genes is there a difference in the distribution of expression levels between the stage II and stage III populations?

This question can be addressed by performing one statistical test per gene, and to define "differentially expressed" genes as those passing some significance threshold.
Below, we use the Wilcoxon rank sum (or Mann-Whitney) test for comparing two independent samples. This is a rank-based non-parametric test. A fast version of this test is implemented in the  `sanssouci::rowWilcoxonTests` function.

To obtain post hoc bounds, we create an object of class `SansSouci` and run the `fit` method to use the (parametric) Simes method introduced by @GS2011:

```{r}
#| label: Simes
obj <- SansSouci(Y = log(1 + Y), groups = groups)
alpha = 0.1
res_Simes <- fit(obj, B = 0, family = "Simes", alpha = alpha,
                 rowTestFUN = rowWilcoxonTests) ## B=0 => no calibration!

```

The target risk level is set to $\alpha = 0.1$, meaning that all of the statements made below will be true with probability greater than $1-\alpha= 0.9$.

# Volcano plots

<!-- For an interactive volcano plot, see the [volcano plot shiny application]( https://shiny-iidea-sanssouci.apps.math.cnrs.fr/). -->

A classical practice in gene expression studies is to define DE genes as those passing both a significance filter (small $p$-value) and an effect size or "fold change" filter. Here, the fold change of a gene is defined as the difference between the expression medians (on the log scale) of the two groups compared. This double selection by $p$-value and fold change corresponds to two sets of genes, with positive/negative fold change, which can be represented in the following plot:

```{r}
#| label: volcano
volcanoPlot(res_Simes, p = 1e-3, r = 1)
```

This type of plot is called a "volcano plot" @CC2003. Post hoc inference makes it possible to obtain statistical guarantees on selections such as the ones represented in the above figure.

# More powerful method

To obtain more informative results, we use the calibration method introduced by @blanchard20post-hoc and @enjalbert-courrech2022powerful. This method uses sample label permutations to learn the gene dependence structure and obtain a more powerful method. 

```{r}
#| label: calibration
#| cache: true
obj <- SansSouci(Y = log(1 + Y), groups = groups)
res_perm <- fit(obj, B = 1000, alpha = alpha, family = "Simes",
           rowTestFUN = rowWilcoxonTests)
```

The annotations in the corresponding volcano plot illustrate a gain in power (more true positives, lower FDP):

```{r}
#| label: volcano_perm
volcanoPlot(res_perm, p = 1e-3, r = 1)
```

# Session information

```{r}
#| label: session-info
sessionInfo()
```

# Reproducibility

To re-build this vignette from its source, use:

```{r}
#| label: reproducibility
#| eval: false
quarto::quarto_render("post-hoc_differential-expression_RNAseq.qmd", output_format = "pdf")
# To keep intermediate files, add option 'clean = FALSE' (rmarkdown-based rendering)
quarto::quarto_render("post-hoc_differential-expression_RNAseq.qmd", output_format = "html")
```

# References
